scRNAseq benchmark in want of a title

### 2019-11-27 bmc_article.tex

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult (Figure \ref{fig:figure3}B). Our method (scDblFinder) achieved the best accuracy while running in a very reasonable time (Figure \ref{fig:figure3}). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure 1D). We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it consistently did (Figure \ref{fig:figure4}).

@@

We next investigated the impact of filtering according to various criteria on the accuracy of the clustering. We tested each aforementioned cell property, alone and in combinations, filtering above and/or below a certain threshold defined by different numbers of median absolute deviations. Reasoning that a bad cell would be bad in several distributions, we also included the optional rule that a cell should be excluded if it is an outlier in at least 2 or 3 distributions.

A first observation was that very stringent filtering (removing up to 20\% of the cells) was not generally associated with an increase in accuracy of the clustering (Supplementary Figure 8A). Moreover, stringent filtering tended to be biased towards the exclusion of specific subpopulations (Supplementary Figure 8B). Indeed, applying the same filtering criteria on individual clusters of cells (identified through scran's \textit{quickCluster} method) resulted in nearly no cell being filtered out, which suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, rather than `bad` cells.

We next investigated the contribution of various criteria individually (Supplementary Figure 9). Interestingly, regression analysis suggests that increasing the number of distributions on which a cell must deviate in order to be excluded tended to be positively associated with ARI, while increasing the number of median absolute deviations required to be an outlier as associated with a decrease in ARI. We therefore suggest excluding cells that are outliers in more than 1 distribution. The impact of filtering on the basis of total counts or feature counts was not always robust, nor always positive, and we could detect no significant impact of filtering on the basis of the proportion of mitochondrial reads - however it must be said that the benchmark datasets count very few cells with a very high proportion of mitochondrial reads (Supplementary Figure 1). We therefore suggest to keep this criterion, but in a relatively lenient fashion, and not solely based on the median absolute deviation of the distribution. Filtering solely based on deviations from the distribution might be inadequate when the distribution is tight; in our experience, the proportion of mitochondrial counts in otherwise `healthy` cells can vary up to 8-12\% depending on the tissue. We therefore recommend using a combination of fixed and deviation based threshold, e.g. x $>$ 0.08 and x $>$ 2.5*MADs.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
    \caption{Proportion of cells filtered out and accuracy of the cluster (under two normalizations) for different sets of filters (see methods).}
    \label{fig:figure4}
\end{figure}

On the basis of these observations, we defined three sets of criteria as lenient, default and stringent filtering (see methods), and investigated them further. We found that filtering had a much milder impact than doublet removal (Figure \ref{fig:figure4}), and that more stringent filtering tended to result in higher silhouette width, in a higher proportion of the variance in the first components explained by subpopulations (Supplementary Figure 10), and often in a higher adjusted Rand index (ARI) of the clustering at the right number of clusters (Figure \ref{fig:figure4}). However, part of the gain is to be ascribed to the removal of cells from subpopulations (especially in the Zheng datasets) that are difficult to distinguish, and for which more stringent clustering did not improve silhouette width (Supplementary Figure 10A). For these reasons, we therefore recommend using a more lenient filter (e.g. lenient or default filter described here) or a cluster-based filter to avoid most of the subpopulation bias while excluding strong outliers.

@@

Mitochondrial reads have been associated with cell degradation, and it has been observed that much of scRNAseq clustering is determined by ribosomal genes [REF]. We therefore considered whether excluding one category of features or the other, or using only protein-coding genes, had impact on the ability to distinguish subpopulations (Supplementary Figure 11). Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations. Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in Seurat, we tested scran's pooling-based normalization \citep{lunPooling2016}, sctransform's variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}. 
In addition to log-normalization, the standard Seurat clustering pipeline scales features so that the PCA is not dominated by a small subset of the features, and we therefore included versions of the different normalization, with or without a subsequent scaling (except sctransform's variance-stabilizing transformation, given that, conceptually, it involves something analogous to scaling). 
Seurat's scaling function includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features. Finally, it has been proposed that the use of stable genes, and in particular cytosolic ribosomal genes, can be used to normalize scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}. We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features, removing the confounding covariation with the subpopulation (Supplementary Figure 12A). The simple step of scaling tended to remove much of the correlation with these features, and in the absence of scaling, standard Seurat normalization resulted in fairly high correlation. sctransform had the lowest correlation, but most methods (including normalization based on stable genes) were able to remove most of the correlation when combined with scaling. The only exception is Seurat normalization with scaling regressing out the number of features, which surprisingly increased the correlation.

We next investigated the impact of normalization on the ability to distinguish subpopulations, which was much greater than that of filtering (Supplementary Figure 13). For many clusters, scaling tended to reduce both the average silhouette width (Supplementary Figure 12B) and the proportion of variance in the top principal components explained by subpopulations (Figure \ref{fig:figure5} and Supplementary Figure 13), and was associated with a poorer adjusted Rand index (ARI) of the resulting clustering (Figure \ref{fig:figure5}). Regressing out covariates systematically gave poorer performance on all metrics. sctransform produced the highest cluster average silhouette width (Supplementary Figure 12B) as well as ARI (Figure \ref{fig:figure5}), but not always the highest mean proportion of variance in the top components explained by subpopulations (Figure \ref{fig:figure5} and Supplementary Figure 13). Of note, even though sctransform is meant to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with smartseq datasets. Interestingly, some clusters tended to be less well distinguished with sctransform than with non-scaled (and in some cases non-normalized) methods, suggesting that some procedures emphasize differences more useful to track some subpopulations than others, and hence the possibility that further developments in methods could improve this step.

@@

Finally, although there were differences in specific cases, we observed no robust tendency of certain normalization methods towards over- or under-estimating the number of clusters when using with Seurat clustering (Supplementary Figure 14).

@@

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations (the latter estimated through edgeR). Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure 14A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 15B-D). We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 16-17). We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Figure \ref{fig:figure6}A). A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure6} and Supplementary Figures 16-17). Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.

@@

We next investigated whether a selection based on these methods impacted the proportion of variance explained by subpopulations in the first principal components (Supplementary Figure 18). To validate the previous assay, included a selection based on the genes whose proportion variance or deviance explained was highest. Interestingly, while these selections were on average the top-ranking methods, they were not systematically so for all dataset.

The main observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and Seurat's un-standardized estimate of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering.

@@

The impact of the dimensionality reduction method was far greater than that of normalization or feature selection (e.g. Supplementary Figure 19). Although the first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations, overall Seurat's PCA procedure proved superior on all metrics (Figure \ref{fig:figure7} and Supplementary Figures 19-21). 
In particular, weighing the components had a strong positive impact on silhouette widths and resulting ARI.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/figure5_dr_files/figure-html/unnamed-chunk-2-1.png}}
    \caption{Evaluation of the dimensionality reduction strategies.}
    \label{fig:figure7}
\end{figure}

>>>>

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the 5 lines dataset (mixology10x5cl) proved more difficult (Figure \ref{fig:figure3}B). Our method (scDblFinder) achieved the best accuracy while running in a very reasonable time (Figure \ref{fig:figure3}). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure 1D). We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it did even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).

@@

We next investigated the impact of filtering according to various criteria (see methods). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g. library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of median absolute deviations (MADs) from the whole distribution can lead to strong biases against certain subpopulations (Figure \label{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \label{fig:figure4}B and Supplementary Figure 8). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here, and we therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm, and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious, and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias. Applying the same filtering criteria on individual clusters of cells (identified through scran's \textit{quickCluster} method) resulted in nearly no cell being filtered out, which suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, rather than `bad` cells. Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy. We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see methods).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in `scater` (see methods for details).}
    \label{fig:figure4}
\end{figure}

@@

Mitochondrial reads have been associated with cell degradation, and it has been observed that much of scRNAseq clustering is determined by ribosomal genes [REF]. We therefore considered whether excluding one category of features or the other, or using only protein-coding genes, had impact on the ability to distinguish subpopulations (Supplementary Figure 9). Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations. Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in Seurat, we tested scran's pooling-based normalization \citep{lunPooling2016}, sctransform's variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}. 
In addition to log-normalization, the standard Seurat clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too dominated by highly-expressed features, and we therefore included versions of the different normalization, with or without a subsequent scaling (sctransform's variance-stabilizing transformation involves something analogous to scaling). 
Seurat's scaling function also includes the option to regress out the effect of certain covariates, and we therefore tested its use with the proportion of mitochondrial counts and the number of detected features. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq; we therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation of sctransform was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}. We therefore first assessed to what extent the first principal component still retained a correlation with the library size and number of detected features, removing the confounding covariation with the subpopulation (Figure \label{fig:figure5}A). The simple step of scaling tended to remove much of the correlation with these features, and in the absence of scaling, standard Seurat normalization resulted in fairly high correlation with technical covariates.  sctransform had the lowest correlation, but most methods (including normalization based on stable genes) were able to remove most of the correlation when combined with scaling. The only exception is Seurat normalization with scaling regressing out the number of features, which surprisingly increased the correlation.

We next investigated the impact of normalization on the separability of the subpopulations (Figure \label{fig:figure5}B-C), and found most methods (including no normalization at all) to perform fairly well. Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width and mutual information. Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones, and was generally but not always beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave poorer performance on all metrics. sctransform systematically outperformed other methods, and even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with smartseq datasets.

@@

Finally, although some normalization methods had a tendency to lead to a higher (e.g. sctransform) or lower (e.g. stable genes) number of clusters, the effect was very mild and not entirely systematic, and we observed no major tendency towards over- or under-estimating the number of clusters (Supplementary Figure 10).

@@

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether each selection method selected genes with a high proportion of variance or deviance explained by (real) subpopulations. Estimates of variance explained by subpopulations based on a standard Seurat normalization or on sctransform data were highly correlated (Supplementary Figure 11A), and there was also a good agreement with deviance explained, although lowly-expressed genes in particular could have a high deviance explained without having much of their variance explained by subpopulations (Supplementary Figure 11B-D). We therefore compared, using both of these estimates, the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13). We first focused on the first 1000 genes to highlight differences between methods, although the differences tended to decrease with a higher number of genes selected (Supplementary Figures 12-14). A first observation was that standardized measures of variability were systematically worse than their non-standardized  counterparts in selecting genes with a high proportion of variance explained by subpopulations, while instead proving most of the time superior in selecting genes with a high deviance explained by subpopulations (Figure \ref{fig:figure6}A and Supplementary Figures 12-13). Deviance proved the method of choice to maximize the former (with mere expression level proving surprisingly good), but surprisingly did not perform so well to select genes with a high deviance explained.

@@

We next investigated whether a selection based on these methods impacted the clustering accuracy (Figure \ref{fig:figure6}A). To validate the previous assay, we included a selection based on the genes whose proportion variance or deviance explained was highest. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all dataset. The previous observations were reflected in the ARI of the resulting clustering (Figure 4B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and un-standardized estimate of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 14).

@@

The impact of the dimensionality reduction method was far greater than that of normalization or feature selection (e.g. Supplementary Figure 15). Although the first components identified by GLM-PCA tended to have a higher proportion of their variance explained by real subpopulations and increase the average silhouette width of already well-defined subpopulations, overall Seurat's PCA procedure proved superior on all metrics (Figure \ref{fig:figure7} and Supplementary Figures 15-17). 
In particular, weighing the components had a strong positive impact on silhouette widths and resulting ARI.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/dr_files/figure-html/unnamed-chunk-2-1.png}}
    \caption{Evaluation of the dimensionality reduction strategies.}
    \label{fig:figure7}
\end{figure}
